---
title: "Average daily activity pattern"
author: "Christian Lindell"
date: "Saturday, July 12, 2014"
output: html_document
---

```{r, warning=FALSE, results='hide', message=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)
# Avoid scientific notation
options(scipen=999)

# Load data
unzip("activity.zip", files = NULL, list = FALSE, 
              overwrite = TRUE,junkpaths = FALSE, exdir = "./data", 
              unzip = "internal", setTimes = FALSE)

data <- read.csv("data/activity.csv")

# Calculate day of week and weekday or weekend
days <- as.POSIXlt(as.character(data$date))
data$day <- days$wday
data$daygrp <- "weekday"
data[data$day==0 | data$day==6, "daygrp"] <- "weekend"
data$daygrp <- as.factor(data$daygrp)

# Impute missing data. Imputed values are set to the mean of the same
# interval in the same day of the week in the whole period, if  f.x interval
# 5 of the first Monday is missing the value are set to the mean of all interval 5 for all Mondays.
# I have choose to not make a different data set for the imputed values but instead just create a 
# new variable in the original data set. The reason to do this is that it makes it easier to 
# compare the imputed value with the original values to see if they are reasonable.

data$impstep <- data$step
for(i in  1:length(data$step)) {
 if (is.na(data$step[i])) {
   data$impstep[i]=mean(data[data$interval==data$interval[i] & 
                               data$day==data$day[i], "steps"], na.rm=T)
 }
}

# Calculate steps per day, excluding NA:s
dateExclNA <- data[complete.cases(data), ]
stepDay <- dateExclNA %>%
  select(date, steps) %>%
  group_by(date) %>%
  summarise(stepsPerDay = sum(steps, na.rm = TRUE))

# Calculate mean steps per day and median steps per day with NA:s excluded
mean(stepDay$stepsPerDay, na.rm=T)
median(stepDay$stepsPerDay, na.rm=T)

# Calculating steps per intervall
stepInterval <- data %>%
  select(interval, steps) %>%
  group_by(interval) %>%
  summarise(stepsPerInt = mean(steps, na.rm=T))

# Find then interval with most steps
maxSteps <- stepInterval[stepInterval$stepsPerInt==max(stepInterval$stepsPerInt), ]

# Calculating steps per day using imputed values for NA:s
stepDayimp <- data %>%
  select(date, impstep) %>%
  group_by(date) %>%
  summarise(stepsPerDay = sum(impstep))

# Calculate average number of steps per interval for weekdays and weekends
stepDaygrp <- data %>%
  select(daygrp, interval, impstep) %>%
  group_by(daygrp,interval) %>%
  summarise(stepsPerInt = mean(impstep, na.rm=T))
stepDaygrp$avgStep <- NA
mean(data[data$interval==5, "impstep"])
for(i in 1:length(stepDaygrp$interval)) {
  stepDaygrp$avgStep[i] <- mean(data[data$interval==stepDaygrp$interval[i], "impstep"])
}
stepDaygrp$diffSteps <- stepDaygrp$stepsPerInt - stepDaygrp$avgStep


totSteps <- data %>%
  select(daygrp, date, impstep) %>%
  group_by(daygrp, date) %>%
  summarise(stepsPerDay = sum(impstep, na.rm=T)) %>%
  group_by(daygrp) %>%
  summarise(stepsPerDaygrp = mean(stepsPerDay, na.rm=T))

totPeriods <- length(data$steps)
totDays <-length(levels(data$date))
startDate <- min(as.character(data$date))
lastDate <- max(as.character(data$date))
missingValues <- sum(is.na(data$steps))
meanStepsPeriod <- mean(data$steps, na.rm = T)
meanImpStepPeriod <- mean(data$impstep)
medianImpStepPeriod <- median(data$impstep)
meanStepsDay <- mean(stepDay$stepsPerDay)
meanImpStepsDay <- mean(stepDayimp$stepsPerDay)
medianStepsDay <- median(stepDay$stepsPerDay)
medianImpStepsDay <- median(stepDayimp$stepsPerDay)
totStepsDayWeekday <- totSteps[totSteps$daygrp=="weekday",2]
totStepsDayWeekend <- totSteps[totSteps$daygrp=="weekend",2]

````



## The mean total number of steps taken per day


The mean steps per day with missing values excluded is `r mean(stepDay$stepsPerDay, na.rm=T)` and the median is `r median(stepDay$stepsPerDay, na.rm=T)`.

  


```{r, message=FALSE}
ggplot(stepDay, aes(x=stepsPerDay)) + 
  geom_histogram() +
  ggtitle("Average number of steps per day") +
  xlab("Number of steps") +
  ylab("Number of days")

```

## The average daily activity pattern

``

```{r}
ggplot(stepInterval, aes(x=interval, y=stepsPerInt)) + 
  geom_line() +
  ggtitle("Average number of steps per 5 minutes interval") +
  xlab("Interval") +
  ylab("Number of steps")

```



The interval that produced the maximum number of steps across all days was `r maxSteps[1,1]` to `r maxSteps[1,1]+5` minutes, that on average produced `r maxSteps[1,2]` steps.


## Imputing missing values


There are a total of `r length(data$steps)` five minutes intervals in the data set,  covering the number of steps over a period of `r length(levels(data$date))` days (from `r min(as.character(data$date))` to `r max(as.character(data$date))`). Of these values a total of `r sum(is.na(data$steps))` contains missing values (coded as NAs). 

The chosen strategy to handle NAs is to impute values by setting the missing values to the average of the same five minutes interval for the same day of the week in the whole dataset, f.x. if the  first interval (0-5 minutes) for the first Monday contains a NA, then it is replaced with the average for the same interval for all Mondays in the data set.

Compared to the original data, the imputed values rising the mean number of steps per interval slightly (from `r mean(data$steps, na.rm = T)` to `r mean(data$impstep)`), but it doesn´t effect the median that is `r median(data$impstep)`. 

The total number of steps per day on average is `r mean(stepDay$stepsPerDay)`. In the data with imputed values the average is `r mean(stepDayimp$stepsPerDay)`. The median value is `r median(stepDay$stepsPerDay)` and the corresponding median for the imputed values  is `r median(stepDayimp$stepsPerDay)`




```{r, message=FALSE}
ggplot(stepDayimp, aes(x=stepsPerDay)) + 
  geom_histogram() +
  ggtitle("Average number of steps per day using imputed values") +
  xlab("Number of steps") +
  ylab("Number of days")

```


## Differences in activity patterns between weekdays and weekends


The number of steps per interval differs depending on if the day is a weekdays or a day in a weekend. As expected the activitys early in the morning are lower in the weekends but after that the number of steps per interval is mostly higher in the weekends. The average number of steps per day in the weekdays is `r totSteps[totSteps$daygrp=="weekday",2]` that can be compared with the average number of steps per day in the weekends, that is `r totSteps[totSteps$daygrp=="weekend",2]`.



```{r}
ggplot(stepDaygrp, aes(x=interval, y=diffSteps)) + 
  geom_line() +
  facet_wrap( ~ daygrp, ncol=1) +
  ggtitle("Average number of steps per five minutes interval in weekdays and weekends, \ndifference from the average number of steps for each interval for all days") +
  xlab("Interval") +
  ylab("Number of steps, differens from average")

```

